---
title: "Investigating car insurance - Final Project"
resource_files:
- .Renviron
output:
  html_notebook:
    code_folding: hide
    toc: yes
    toc_depth: 4
    toc_float: yes
runtime: shiny
---

```{r setup, include=FALSE}
library(tidyverse)
library(data.world)
library(shiny)
knitr::opts_chunk$set(echo = TRUE)
```

## Github Link

Click [here to view Team 3's Github](https://github.com/omar-olivarez/F17eDVProject4).

## Data.world Project Link
Click [here to view Team 3's Data.world Project Link](https://data.world/omarolivarez/f-17-edv-project-4).

## What are we working on? An introduction to our project

This notebook contains the different steps in the Data Science Pipeline, which describes how data is imported, cleaned, exported, reformated, transformed, visualized, and communicated. Each step will be it's own section in this notebook, with the whole of this notebook serving as an example of communication.  

## What's this data? A description of the US Crimes dataset

The dataset used for all these examples is data gathered by the FBI on criminal activities between the years 1994 to 2013. The data contains the total number of crimes recorded for that year and also the rate of the crime per 100000 people. 

Crimes included are: Violent crimes, Murder, Rape, Robbery, Aggrevated assualt, Property crimes, Burglary, Larceny-Theft, Motor Vehicle Theft.

## Inputting the data and cleaning it 
Let's first import the dataset as a csv from the hard disk and clean the columns so that they are in integers and decimals rather than characters. Click on the 'Code' button to see that we used the 'read_csv' function to read the 'Crime_US_94-13.csv' file from the hard disk of the user 'OmarOlivarez'.
```{r}
df <- read_csv("Crime_US_94-13.csv", col_types = list(
  year = col_number(),
  pop = col_number(),
  violent_crime = col_number(),
  pop = col_number(),
  murder = col_number(),
  murder_rate = col_number(),
  rape = col_number(),
  rape_rate = col_number(),
  robbery = col_number(),
  robbery_rate = col_number(),
  agg_assault = col_number(),
  aggravated_assault_rate = col_number(),
  property_crime = col_number(),
  property_crime_rate = col_number(),
  burglary = col_number(),
  burglary_rate = col_number(),
  larceny_theft = col_number(),
  larceny_theft_rate = col_number(),
  motor_vehicle_theft = col_number()
)) 
for(n in names(df)) {
  df[n] <- data.frame(lapply(df[n], gsub, pattern="[^ -~]", replacement= ""))
}
write_csv(df, "cleanedCrimeData.csv")
```
Now the columns in the dataframe are either integers (for example: year) or decimals (for example: murder_rate). Click on the 'Code' button above. You can see that our team exported the new file 'cleanedCrimeData.csv' to the hard disk of the same user 'OmarOlivarez'. After this step, our team loaded the cleaned csv into data.world for the proceeding steps in the Data Science Pipeline process.

## Importing cleaned data from data.world
Now let's use the data.world library to import the [US Crimes dataset](https://data.world/omarolivarez/crime-in-the-united-states) from data.world:
```{r}
project <- "https://data.world/omarolivarez/f-17-edv-project-4"
data.world::set_config(cfg_env("DW_API"))
df <- data.world::query(data.world::qry_sql("SELECT * FROM cleanedcrimedata"), dataset = project) %>% dplyr::select(1:19)
```
Let us take a look at the dataset's column names to see which visualizations work best for the data types:
```{r}
names(df)
```
## Crime Rates Decrease, But Proportions Remain the Same (Line Graph)

### Changing the Data We Have to Compare Crime Rates to Crime Proportions (Reformat,Transform, & Visualize)
Click on the 'Code' button below to see our code. First, we selected the columns for crime rates per 100,000 people and mutated them to be decimals from 0 to 1. 

Next, we took the columns that had the absolute number of occurrences for each type of crime, and divided each by the total number of crimes in that year to get each crime's proportion of occurrences per year.

We overlayed these two newly mutated columns to be able to compare crime rates over time to proportions of crimes over time. Below we can see our findings. **It can be observed that although crime rates in general seem to demonstrate a downward trend over the 20 year period, that crime proportions stay relatively stable over time.**
*(Solid lines depict crime rates while dashed lines depict crime proportions)*
```{r}
df2 <- df%>% dplyr::mutate(violent_crime_rate = violent_crime_rate / 100000, murder_rate = murder_rate / 100000, rape_rate = rape_rate / 100000, robbery_rate = robbery_rate / 100000, aggravated_assault_rate = aggravated_assault_rate / 100000, property_crime_rate = property_crime_rate / 100000, burglary_rate = burglary_rate / 100000, larceny_theft_rate = larceny_theft_rate / 100000, motor_vehicle_theft_rate = motor_vehicle_theft / pop) %>%tidyr::gather(violent_crime_rate, murder_rate, rape_rate, aggravated_assault_rate, property_crime_rate, burglary_rate, larceny_theft_rate, motor_vehicle_theft_rate, key = "crime", value = "rate_per_capita")

df3 <- df%>%dplyr::mutate(prop_murder = murder / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_rape = rape / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_robbery = robbery / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_agg_assault = agg_assault / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_property_crime = property_crime / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_burglary = burglary / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_larceny_theft = larceny_theft / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft), prop_motor_vehicle_theft = motor_vehicle_theft / (murder+rape+robbery+agg_assault+property_crime+burglary+larceny_theft+motor_vehicle_theft)) %>% dplyr::select(year, prop_murder:prop_motor_vehicle_theft) %>% tidyr::gather(prop_murder:prop_motor_vehicle_theft, key = 'crime_type', value = 'prop_total_crime')

myColors <- c("red4", "darkslategray3", "dodgerblue1", "darkcyan","gray79", "black", "skyblue2", "dodgerblue4","purple4", "maroon", "chocolate1", "bisque3", "bisque","seagreen4", "lightgreen", "skyblue4", "mediumpurple3","palevioletred1", "lightsalmon4", "darkgoldenrod1")

renderPlot({ 
  p <- ggplot()
  p <- p + geom_smooth(data = df2, mapping = aes(x= year,y=rate_per_capita, colour = crime), size = 2)
  p <- p + geom_smooth(data=df3, mapping = aes(x= year,y = prop_total_crime/4, colour = crime_type),  linetype = "dashed", size = 2) + scale_color_manual(values=myColors)
  p <- p + scale_y_continuous(sec.axis = sec_axis(~.*4, name = "Proportion out of total crime"))
  p
  })
```

## Cumulative Distribution of Crimes
In the following example, we used gather, group by, cume_dist, arrange, summarise, and mutate to evaluate the cumulative distribution of crimes. See the table below on how crimes rank from highest to lowest. Click on the 'code' button to see our code.

```{r}
df5 <- df %>% dplyr::select(violent_crime, murder, rape, robbery, agg_assault, property_crime, burglary, larceny_theft, motor_vehicle_theft) %>% tidyr::gather(violent_crime:motor_vehicle_theft, key = "crime", value = "count") %>% dplyr::group_by(crime) %>% summarise(count = sum(count)) %>% dplyr::arrange(desc(count)) %>% dplyr::mutate(crime_percent = cume_dist(count))
shiny::renderDataTable(df5)
```


### 20 Years at a Glance: Diminished Crime Rates from 1993 to 2013 
The bar chart below shows the different crime rate based on the year. To get the specific colums from the csv file, select is needed combined with the input panel.

```{r}
crimes <- c("violent_crime_rate","murder_rate","rape_rate","robbery_rate","aggravated_assault_rate","property_crime_rate","burglary_rate","larceny_theft_rate")

inputPanel(
  selectInput("selectCrime", label = "Crime:",
              choices = crimes, multiple=FALSE, selected="violent_crime_rate")
)

renderPlot({ 
  kpdf2 <- df %>% dplyr::select(year,input$selectCrime) 
  kpdf2 %>% ggplot() + geom_col(mapping = aes(x=year,y= kpdf2[2])) + scale_y_continuous(name =input$selectCrime)
  })
```

### Percent rank of violent crime throughout the years. 

We can look at what year hosted the highest percentage of crimes of violent crimes in the data. 

```{r}
pcnt_violent = df %>% dplyr::select(year, pop, violent_crime_rate) %>% dplyr::mutate(prcnt = percent_rank(violent_crime_rate)) %>% dplyr::arrange(year)

shiny::renderDataTable(pcnt_violent)
```

As we can see there has been a decreasing tredn in violent crimes even when population has been stedily increasing. 

## PieChart of Violent Crimes
The pie chart below is an example of mutate to get the perecent for total violent crime. 

```{r}
kpdf1 <- data.world::query(data.world::qry_sql("SELECT year, pop, violent_crime FROM cleanedcrimedata"), dataset = project) %>% dplyr::mutate(violent_crime_per_capita = violent_crime/pop) %>% dplyr::mutate(per_total_per_capita = violent_crime_per_capita/sum(violent_crime_per_capita))

kpdf1$label <- scales::percent(kpdf1$per_total_per_capita)

renderPlot({
 ggplot(kpdf1)+
  geom_bar(aes(x="", y=per_total_per_capita, fill=year), stat="identity", width = 1)+
  coord_polar("y", start=0) + theme_void() + 
  geom_text(aes(x=1.35, y = cumsum(per_total_per_capita) - per_total_per_capita/2, label=label))
})

```


### Change in different crimes across the years.

Using lag we can calculate the change in difference of crime throughout the years. 

```{r}
crime_change = df %>% dplyr::select(year, pop, robbery_rate, burglary_rate, larceny_theft_rate) %>% 
  dplyr::mutate(delta_robbery_rate = robbery_rate - lag(robbery_rate), delta_burglary_rate = burglary_rate - lag(burglary_rate), delta_larceny_theft = larceny_theft_rate - lag(larceny_theft_rate))

shiny::renderDataTable(crime_change)

crime_changevz = crime_change %>% dplyr::select(year, delta_robbery_rate, delta_burglary_rate, delta_larceny_theft) %>% tidyr::gather(delta_robbery_rate:delta_larceny_theft, key = "crime_type", value = "delta") %>% dplyr::arrange(year)

renderPlot({
  ggplot(crime_changevz, aes(x = year, y = delta, color = crime_type)) + geom_line()
})
```

As we can see the changes in crime rates follow each other, with robbery having the highest amount of changes while larcerny and theft have the lowest changes. 